---
title: "Machine Learning for Human Activity Recognition: Weight Lifting Exercises"
author: "Sviataslau Kohut"
date: "March 18, 2016"
output: html_document
theme: default
---
```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(caret)
library(corrplot)
set.seed(12345)
```

## Introduction
Using devices such as *Jawbone Up*, *Nike FuelBand*, and *Fitbit* it is now possible to collect a large amount of data about personal activity relatively inexpensively. 
These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. 
One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. 
In this project, data from accelerometers on the belt, forearm, arm and dumbell are used to predict quality of execution of dumbell lifts using the random forest algorithm. 
A model that was constructed involves the use of 46 predictors and allows one to predict the execution quality with the accuracy of 99.42 %. 

## Exploratoty Data Analysis

A training data set contains around 20 000 observations for 160 variables, while there are only 20 of them in a testing set.

```{r, cache=TRUE}
trainingSet<-read.csv('/home/molkee/Desktop/8-practical-machine-learning/pml-training.csv',sep = ',',dec='.')
testingSet<-read.csv('/home/molkee/Desktop/8-practical-machine-learning/pml-testing.csv',sep = ',',dec='.')
dim(trainingSet)
dim(testingSet)
```

A variable to be predicted is the *classe* variable. It has five values: A, B, C, D and E:

```{r,echo=FALSE}
levels(trainingSet$classe)
```

The class "A" refers to execution exactly as in the specification, while other classes  represent common execution mistakes.

We need to get rid of the irrelevant data such as time stamps, participant name, etc. These data are the first seven columns in the data sets:  

```{r}
str(trainingSet[,c(1:7)])
trainingSet<-trainingSet[,-c(1:7)]
testingSet<-testingSet[,-c(1:7)]
```

Because training data set now only contains response from accelerometers (predictors) (*numeric*","*integer*" and "*factor*"), we need to convert all *factor* predictor variables to *numeric* as they clearly represent numerical data (acceleration, coordinates, etc.).

```{r,warning=FALSE}
unique(sapply(trainingSet[,-ncol(trainingSet)],class))
# the same indices in case of the testing set
indFactor<-which(sapply(trainingSet[,-ncol(trainingSet)],is.factor))
trainingSet[indFactor] <- lapply(trainingSet[indFactor], function(x) as.numeric(as.character(x)))
testingSet[indFactor] <- lapply(testingSet[indFactor], function(x) as.numeric(as.character(x)))
```

Finally, all *NA* values were set to zero. 

```{r}
trainingSet[is.na(trainingSet)]<-0.
testingSet[is.na(testingSet)]<-0.
```


## Pre-Processing

Predictors in the data set need to be preprocessed.

The plan is as follows:

1. eliminate zero and near-zero variance predictors
2. eliminate correlated predictors  
 
First, near-zero variance predictors were eliminated using the `nearZeroVar` function:
```{r,warning=FALSE}
nzv<-nearZeroVar(trainingSet[,-ncol(trainingSet)],freqCut = 95/5,uniqueCut=10)
trainingSet<-trainingSet[,-nzv]
```

Next, the `findCorrelation` function was applied to remove predictors with absolute correlations above 0.9:
```{r,warning=FALSE}
corPred<-findCorrelation(cor(trainingSet[,-ncol(trainingSet)],use='complete.obs'),cutoff = 0.90,verbose = F)
trainingSet<-trainingSet[,-corPred]
```

The resulting data clean set contains 45 predictors and the target `classe` variable.
```{r}
dim(trainingSet)
```

## Model Training and Tuning 

The data set was partiotioned into a training set (80 % of the data) and a validation set (20 % of the data) which was used to access quality of fits. 
```{r}
indTrain<-createDataPartition(trainingSet$classe, p = .8, list = FALSE)
training<-trainingSet[indTrain,]
validation<-trainingSet[-indTrain,]
```

Several models were considered for predicting the quality of execution.  
They included decision trees (`rpart`), stochastic gradient boosting (`gbm`) and the random forest algorithm (`rf`).
The latter is vastly superior to the other models in terms of accuracy and computational.
Therefore, the random forest algorithm was applied to the problem under consideration. 
Naive application of the `train` function in conjunction with the `rf` method for the training data set is computationally expensive and may require several hours.
To speed up the code, some parameters were tuned. 
In particular, parallelization was used (`doMC` library). 
The code was executed on two cores.
Also, the `mtry` parameter (a number of predictors at each node of the tree) was set to 10. 
This made it possibe to reduce computational time to less than half an hour.
Repeated K-gold (*K*=5) cross validation was used for better accuracy. 
```{r fit,cache=TRUE, warning=FALSE}
library(doMC)
registerDoMC(2) #use two cores

fitParam<-trainControl(method ="repeatedcv",
                           number = 10,
                           repeats = 5,
                           returnResamp = "all",
                           allowParallel = T)
rfGrid<-expand.grid(mtry = 10)

rfFit <- train(classe ~ ., 
              data = training,
              method = "rf",
              trControl = fitParam,
              rfGrid = rfGrid,
              metric="Accuracy",
              maximize=TRUE)

rfFit
```



## Model Assessment

To access performance of the prediction model, it was applied to the validation set.
The resuting prediction accuracy is 99.41 %.
This is quite impressive because the accuracy achieved on the training set was 99.42 %.
```{r pred-valid,cache=TRUE, dependson='fit',warning=FALSE}
cM<-confusionMatrix(validation$classe,predict(rfFit,newdata=validation))
cM
```

## Prediction

The obtained model was used for predictions on the testing set.  
It accurately predicted *all* outcomes for twenty provided observations.  
```{r pred-test,cache=TRUE, dependson='fit',warning=FALSE}
results<-predict(rfFit,newdata=testingSet[-ncol(testingSet)])  
results
```

## Appendix 

1. **Confusion Matrix Heatmap**

```{r fig1}
my_scale = c(0, 10, 100, 1000, 10000)
melted_cM<-as.data.frame.table(cM$table)
ggplot(melted_cM)+
geom_tile(aes(x=Prediction, y=Reference, fill=Freq))+
scale_x_discrete(name="Actual Class")+
scale_y_discrete(name="Predicted Class")+
scale_fill_gradient(name="freq", trans="log",low="red",breaks=my_scale)+
theme_bw()+
ggtitle("Confusion Matrix")
```

2. **Correlation Matrix for Predictors**

```{r fig2,fig.width=8,fig.height=8}
corrplot(cor(training[-ncol(training)]),method="color")  
```
